---
title: "Solar Panels at NAIT"
author: "Alexander Ondrus"
date: "07/03/2020"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(tcltk)
solarFilename <- "Historical_-_City_of_Edmonton_NAIT_Solar_Panel_Reference_System.csv"
if ( !file.exists(solarFilename) )
{ 
  msg <- paste(
    "Missing CSV data. ",
    "Click \'OK\' to download to working directory, \'Cancel\' to skip.",
    sep="")
  msgBox <- tk_messageBox(title = "Waiting for download...",
                          message = msg, icon = "info", type = "okcancel")
  if( identical(msgBox, "ok") )
  {
    download.file(
      "https://data.edmonton.ca/api/views/uwuq-3zq3/rows.csv?accessType=DOWNLOAD",
      destfile = solarFilename,
      method = "auto"
    )
#    browseURL("https://data.edmonton.ca/Environmental-Services/Historical-City-of-Edmonton-NAIT-Solar-Panel-Refer/uwuq-3zq3")
  }
}
if ( !file.exists(solarFilename) )
{
  msg <- paste(
    "CSV data file not detected.",
    "Please ensure working directory is set correctly before proceeding. ",
    "It is currently: \n",
    getwd(),
    sep=""
  )
  msgBox <- tk_messageBox(title = "Error", message = msg, icon = "info", type = "ok")
}
```

## Loading the Data

First, I load the `tidyverse` library and read the downloaded CSV file using the `read_csv()` command. As long as the csv is stored in the same directory as this file, you can use tab-autocomplete to find the file quickly.

The notation `<-` to store the result of a command in another variable can be used interchangably with `=`.

```{r Load file, message=FALSE, warning=FALSE}
library(tidyverse)
solar <- read_csv("Historical_-_City_of_Edmonton_NAIT_Solar_Panel_Reference_System.csv")
```

Data is stored internally as a _data frame_. Click on the view icon on the right side of the 'Environment' pane to see the first few rows. Note that R automatically recognized the date column as dates and the time column as times, and so we can perform calculations on them and integrate the results in the text. 

To access columns in a data frame we use the `$` notation, and we can perform functions such as `max()`, `min()`, `mean()`, etc. on the column in a single command.

```{r Find earliest and latest dates}
min_date <- min(solar$DATE)
max_date <- max(solar$DATE)
```

To use these results integrated into the text, we use single tic-marks as follows; the data begins on `r min_date` and ends on `r max_date`. When we knit the document the variable names are replaced by the results of the calculations. This is useful when generating reports that have external data references. The text then automatically updates with the data source.

## Data Wrangling - Combining, Selecting, Pivoting, Filtering, Aggregating

### Combining

Notice that the date and time is currently contained in two separate columns. The `lubridate` package contains tools to easily combine our date and time columns into a single 'date-time' object, which we can then use for our x-axis in a plot later on.

To add the two together using `lubridate`, I literally add the two columns (first using the functions `ymd()` and `hms()` to have the two columns recognized as year-month-day and hour-minute-second, respectively) 

```{r Combining Date and Time, message=FALSE, warning=FALSE}
library(lubridate)
solar$DATE_TIME <- ymd(solar$DATE) + hms(solar$TIME)
```

### Selecting

For simplicity, I am going to look at only the columns that deal with power. I can pick out specific columns using the `select()` function. Note that all of the columns that I am interested in (excluding the `DATE_TIME` column) end in `POWER`.  The `tidyselect` package (part of the `tidyverse` that we already loaded) has convenient function for just this occasion: `ends_with()`.

```{r Selecting Date Time and Power columns}
power <- select(solar, DATE_TIME, DATE, ends_with("POWER"))
```

### Pivoting

R works best with [tidy data](https://en.wikipedia.org/wiki/Tidy_data). In our case, the data is not "tidy" because the column names contain some of the data (the angles of the various solar panels).

To tidy the data, I want to gather the data in the column headers in its own column. This is done by `gather`ing the columns into more rows using the `gather()` function. This is sometimes known as _long_ format, because it increases the number of rows while reducing the number of columns.

```{r Pivot to long format}
power$DATE_TIME <- as.POSIXct(power$DATE_TIME)
power_long <- gather(power, key = "Angle", value = "Power", -DATE_TIME, - DATE)
```

I am going to clean up the `Angle` column by removing the characters "_" and "POWER" from each of the entries. This can be done more elegantly using regular expressions, but for simplicity I just do a direct character replacement.

```{r Clean Angle Columns}
power_long$Angle <- gsub("_POWER", "", power_long$Angle, fixed = TRUE)
power_long$Angle <- gsub("_", " ", power_long$Angle, fixed = TRUE)
```

### Filtering

Notice that there are `r sum(is.na(power_long$Power))` entries in the `Power` column that have values of `NA`. This is R's way of indicating that a value is missing. I don't want to include these values in my plots below, so I am going to remove them using the `filter()` command. Similar to Python, the exclamation point (!) is used to indicate 'not'.

```{r Filtering out NA values}
power_long <- filter(power_long, !is.na(Power))
```

### Aggregating

There are too many points to make a meaningful plot of all of the data, so I want to take the maximum power values on each day. To do this I will use `group_by()` to set the levels that I want R to take for my groups, then I will use the `summarise()` function to summarise over each of those groups. I connect the two functions with the _pipe_ operator ` %>% `, which takes the output of the first and feeds it to the second.

```{r Daily max power}
power_daily <- group_by(power_long, DATE, Angle) %>% 
  summarise(Max_Power = max(Power))
```


## Visualizing

### Question 1 - How does power depend on the date?

The first way that I want to visualize the data is to plot the power on the y-axis, the date on the x-axis, and colour the points by angle. I will generate this (and all of the other plots today) using the `ggplot2` package. The `ggplot2` package is part of the `tidyverse` and so it has already been loaded. The basics of how to generate plots in `ggplot2` are:

1. Plots are generated by starting with a `ggplot()` command which contains the data source and the `aes`thetics (mappings from data to visual elements) that will be used on all subsequent layers (by default, but this can be changed).
2. Layers are added to the plot by using the `+` symbol.
3. Plots are stored to a variable and then plotted using the `plot()` command or saved to an external file using the `ggsave()` command.

In my case, the data is `power_daily` and the `aes`thetics I specify are `x = DATE`, `y = Max_Power`, and `colour = Angle`. The layers we will need are `geom_point()` (for points), `geom_smooth()` (to add smoothing curves), and `labs()` for labels.

```{r Scatter plot time vs power, message=FALSE}
power_daily$Angle <- factor(power_daily$Angle,
                            levels = c("90 RIGHT",
                                       "53 RIGHT",
                                       "45 RIGHT",
                                       "27 RIGHT",
                                       "18 RIGHT",
                                       "14 RIGHT",
                                       "14 LEFT",
                                       "18 LEFT",
                                       "27 LEFT",
                                       "45 LEFT",
                                       "53 LEFT",
                                       "90 LEFT"),
                            ordered = TRUE)

p <- ggplot(power_daily, aes(x = DATE, y = Max_Power, colour = Angle)) +
  geom_point(alpha = 0.1) + 
  geom_smooth(se = FALSE) +
  labs(title = "Power Output of Solar Panels on NAIT Roof",
       subtitle = "PV panels installed on roof of Shaw Theatre, Main Campus",
       x = NULL,
       y = "Maximum Daily Power",
       colour = "Angle of\nSolar Panel",
       caption = "Data: City of Edmonton Open Data") +
  theme_minimal()
plot(p)
```

This broadly answers the first question. The two trends that I notice are:

- The panels that are _not_ angled at $90^0$ have peaks in the summer and troughs in the winter
- The panels that _are_ angled at $90^0$ have the opposite trend

### Question 2 - Which angle produces the best power?

I am not interested in the power production for a specific day, but the _distribution_ of the power production for the cells installed at different angles. This means that my `x` aesthetic will be the max power, and I will choose my `fill` aesthetic to be my angle so that I can differentiate between them. I'll make the base plot and add some labels, storing the object as `q`. 

```{r Distribution of powers base plot}
q <- ggplot(power_daily, aes(x = Max_Power, fill = Angle)) +
  labs(title = "Which Angle Has the Highest Power Output?",
       subtitle = "PV cells installed on the roof of NAIT Main Campus",
       caption = "Data: City of Edmonton Open Data",
       x = "Maximum Daily Power",
       fill = "Angle of\nInstallation")
```

The first way that I will try and get a sense for the distribution is to use the `geom_density` to build _probability density distributions_ for each angle.

```{r Overlapping density curves}
q1 <- q + geom_density(alpha = 0.5) +
  labs(y = NULL)
plot(q1)
```

Even using the opacity to account for the overlap, it is difficult to distinguish the curves. In order to make that easier, I will generate a [ridge plot](https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html). Notice that the order of the angles has already been set using the `factor()` command in the last plot. This requires us to load (after installing) the `ggridges` package.

```{r Ridge plot using ggridges, message=FALSE, warning=FALSE}
library(ggridges)

q2 <- q + geom_density_ridges(aes(y = Angle),
                              quantile_lines = TRUE,
                              quantiles = 2) +
  guides(fill = "none")
plot(q2)
```

It looks like $53^0$ left is the winner! Just to double-check, I will calculate the median values for each of the distributions and make a simple bar graph.

```{r Calculate median for each angle}
medians <- group_by(power_daily, Angle) %>% 
  summarise(`Median Power` = median(Max_Power))

median_bars <- ggplot(medians, aes(x = Angle, y = `Median Power`, fill = Angle)) +
  labs(title = "Median Maximum Daily Power",
       subtitle = "PV cells installed on the roof of NAIT Main Campus",
       x = "Angle of Installation",
       caption = "Data: City of Edmonton Open Data") +
  geom_col() + 
  guides(fill = "none") +
  geom_label(aes(label = `Median Power`)) +
  coord_flip()

plot(median_bars)
```

